---
title: "Project Proposal"
author: "Broderick Prows"
date: "10/28/2020"
output: pdf_document
---


End Goal of Project: Can we predict Off-Speed Pitches? If yes, is there a succint way (summary statistics) to remember what pitch is most liekly when you are in the box? 


Here is an example of the Baseball Savant URL for Padres opening day 2019. This is where we can get pitch velocity and pitch types, as well as high-level results (ball,strike,inplay,outs). This part will take a bit of work to get a viable webscraper to extract the data. 
```{r}
opening_day_url <- "https://baseballsavant.mlb.com/gamefeed?game_pk=566275&game_date=2019-3-28&type=pitch_velocity&chart_view=pitch&chart_type=callName&inning=&count=&batter_hand=&pitcher_hand=&filter=&player=home-pitchers_489334&view=Umpire&coloring=Pitch%20Type"
```

The situational data can be extracted from [RetroSheet Event](https://www.retrosheet.org/eventfile.htm) files. Worst comes to worst we can try to finesse a project out of this data. (predict HR ro something). 
```{r}

```

We will probably want to pull another file to set up prior distributions for pitch types, or leanr that from the first part of 2019 season. 
```{r}
#require(devtools)
#install_github("BillPetti/baseballr")
require(baseballr)
```

Once we can get this data organized, might want to consider storing it in MySQL, depending on how long and robust the process can be.



N ~ Poisson(mu)
Y|N ~ Binomial(N, p)
Y1
Y2|Y1
Y3|Y2, Y1
P(Y1, Y2, …, YN, N)
P(Y1, Y2, …, YN, N)=P(N)P(Y1|N)P(Y2|Y1, N).....

Mixture of multinomial and poisson, auto-regressive


```{r}
playerid_lookup(last_name = "Scherzer", first_name = "Max")
```

Lance Lynn another good edge case.
```{r,message=F}
pitcher <- 
  scrape_statcast_savant(
    start_date = "2015-03-25", 
    end_date = "2015-10-01", 
    playerid = 453286,  
    player_type='pitcher')
prop.table(table(pitcher$pitch_type))
```


```{r}
bauer_outage <- 545333
last_bauer <- scrape_statcast_savant(start_date = "2016-04-24", end_date = "2020-04-24", playerid = bauer_outage, player_type='pitcher')
```

```{r}
tail(last_bauer)
```

```{r}
hendricks <- scrape_statcast_savant_pitcher(start_date = "2016-04-24", end_date = "2016-05-01",pitcherid = 543294)
```

